---
title: "Data_extractions"
author: "ML"
date: "10 January 2019"
output: html_document
---


```{r setup, cache = F, echo=FALSE}
knitr::opts_chunk$set(error = TRUE) #allow some execution errors for demonstration purposes
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE)
sessionInfo()
```

# EXERCISES - PHYLOGENETIC TREES FOR META-ANALYSIS 

Prepared by Malgorzata (Losia) Lagisz, BEES, UNSW, AU, losialagisz@gmail.com.   

### Purpose of the exercise:   
In this exercise you get some practice and tips on:   
 * Organising dataset  
 * Extracting and coding
 * Converting different types of information   
 * Tidying up

  

## Introduction  
-------

Imagine, you decided to carry on your Terminal Investment Meta-analysis and you actually fnished searching and screening literature for relevant papers. You have a collection of around 30 include full-text papers stored neatly as pdf files and supplementary information in whatever format available.



********************************************************************************
###  Organising dataset  


********************************************************************************
###  Extracting and coding

**TO DO**:  
Prepare a set of 3 DAH papers to be extracted using *metaDigitise* package
See: https://github.com/daniel1noble/metaDigitise (link to the paper?)
Tools: metaDigitise


********************************************************************************
###  Converting different types of information   

**TO DO**:  
ES_calculator to convert ES?

********************************************************************************
###  Tidying up






Start from installing (as required) and loading R packages.   

```{r prepare, message = FALSE}
install.packages("devtools")
devtools::install_github("daniel1noble/metaDigitise")
#install.packages("metaDigitise")
library(metaDigitise)
```
